Wormtable: A data structure for genome scale data

### 2013-07-04 docs/paper/paper.tex
Suggestions + edits to examples section

% but we use text files
Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are almost entirely 
based around the use of text files. Plain text files have many advantages:
they are extremely portable and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.

@@

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and avoiding the need
for repeated parsing. Records are usually stored in a
B-tree, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

@@

@@

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of a relational database 
server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store~\citep{m12}. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads of a server process

@@

% What is VCF?
A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In VCF,  information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11}, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format using the included \texttt{vcf2wt} 
conversion program.  The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows. Methods using the VCF directly must parse 
each row before it can be used, and this parsing is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[citation?]. Both of these
methods proceeded row-by-row, checking the ALT and REF columns. The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

It is well known, of course, that plain text cannot be retrieved 
efficiently at large scales. The BVCF format is an binary version 
of VCF which is intended to avoid this overhead [citation?]. 
This binary format does alleviate the problem of parsing overhead,
but does not solve the problem of indexing \emph{within} rows.

To illustrate the use of indexes in wormtable, 
we created an index on the ALT and REF columns, 
This index required X MB of disc space Y minutes to build, respectively. 
The ALT+REF index provides a much more efficient means of counting 
the transitions and transversions, and another script to calculate 
these required x seconds.

>>>>

% but we use text files
Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are almost entirely 
based around the use of text files. Plain text files have many advantages:
they are extremely portable, easy to view and interpret and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.

@@

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which can solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and avoiding the need
for repeated parsing. Records are usually stored in a
B-tree, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

% there are a range of standard (well defined) tabular data formats: vcf, gtf
% lends itself to a database solution geared towards specific file types

@@

% are other database solutions less flexible when it comes to where the 
% data is stored?

@@

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of a relational database 
server, wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store~\citep{m12}. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. DB is a highly modular library, providing the  % do you mean Berkeley DB?
means to define a very lightweight read-only database, without the 
overheads of a server process

@@

% What is VCF?
A popular file format for Bioinformatics use is the Variant Call Format~\citep{da+11},
or VCF. In a VCF file, information about variant sites in a genome is encoded 
as tab-delimited rows in a text file. To access information about a 
particular site naively, one must proceed through the file line-by-line until the 
desired record is found. There are methods available to index this 
file~\citep{li11} by genomic position, but these have no semantic 
understanding of the structure of the file and so are limited in the 
fields that can be indexed. One cannot, for example, build an index 
to quickly obtain all records in which the quality field is greater
than some value. Such queries can take a considerable time on large VCF 
files, which discourages testing and checking for the effects of 
arbitrary thresholds in analyses.

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
wormtable format using the included \texttt{vcf2wt} conversion program.  
The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
which required X hours.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows whereas methods that use the VCF directly 
must parse each row before it can be used, and is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[citation?], a module for
python to parse VCF files. In both of these examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
summing transitions and transversions (ignoring indels). The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

It is well known, of course, that plain text cannot be retrieved 
efficiently at large scales. The BVCF format is a binary version 
of VCF which is intended to avoid this overhead [citation?]. 
This binary format does alleviate the problem of parsing overhead,
but does not solve the problem of indexing \emph{within} rows.

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns. This index required X MB of disc space 
Y minutes to build, respectively. The REF+ALT index provides a much more 
efficient means of counting the number of transitions and transversions 
(we can just count the number of rows matching keys in the index 
corresponding the SNPs of each type). A script to calculate these 
counts by this method required x seconds.
